Individual and Differential Harm in Redistricting

### 2025-08-25 paper/harm.qmd
discsussion sentence
edits

Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, Massachusetts' 2010 districts had an efficiency gap that would normally qualify as a Democratic gerrymander, but was in fact due to the homogeneity of the state's political geography [@duch:18].
While consideration of counterfactual plans is now commonplace with redistricting simulation algorithms [e.g., @deford2019], our approach is different in that it incorporates them within the definition of the measure itself.
As far as districting metrics themselves, many rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
By making the comparison set explicit, our approach improves on existing work by better reflecting legal and geographic realities, while making the interpretation of our measures clearer.

@@

A second limitation is that the harm definition ignores any preferences a voter may have between the candidates that they did not vote for.
For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
The simple cost function implicitly assumed in our proposed measure does not differentiate between candidates B or C winning the district.
The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.

@@

>>>>

Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, Massachusetts' 2010 districts had an efficiency gap that would normally qualify as a Democratic gerrymander, but was in fact due to the homogeneity of the state's political geography [@duch:18].
While consideration of counterfactual plans is now commonplace with redistricting simulation algorithms [e.g., @deford2019], our approach is different in that it incorporates them within the definition of the measure itself.
Some existing districting rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
By making the comparison set explicit, our approach improves on existing work by better reflecting legal and geographic realities, while making the interpretation of our measures clearer.

@@

A second limitation is that the harm definition ignores any preferences a voter may have between the candidates that they did not vote for.
For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
The simple cost function implicitly assumed in our proposed measure does not differentiate between candidates B or C winning the district.
The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
For example, in our Boston example, the distance in latent space between a voter's ideal point and the winning candidate's position could be used.

@@

One implication of the model presented here is that it rules out any mechanism whereby voters reward or punish the parties based on the plan enacted: $q$ itself does not show up in the model.
Indirect effects, such as $q$ affecting the type of candidates who run, and therefore the vote share each party wins, are also ruled out, or at least on average: the residual $\varepsilon_{dv}$ captures candidate effects broadly, but it is assumed to be drawn independently of $q$.
To the extent that a particular $q$ affects voter turnout or vote choice in these ways, the simple model presented here will not capture those effects.
